---
title: "World Cup Model Analysis"
author: "Edgar Gonzalez Martinez"
date: today
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(nnet)

if (!requireNamespace("glmnet", quietly = TRUE)) {
  stop("Package 'glmnet' is required. Install it with install.packages('glmnet').")
}

set.seed(131)
```

# Load Data

```{r}
world_cup <- read_csv(
  "data/world_cup_pre_tournament_predictors.csv",
  show_col_types = FALSE
)

train_data <- world_cup |>
  filter(split == "Training") |>
  mutate(
    match_result = factor(
      match_result,
      levels = c("Home Win", "Draw", "Away Win")
    ),
    home_confederation = factor(home_confederation),
    away_confederation = factor(away_confederation)
  )

test_data <- world_cup |>
  filter(split == "Testing") |>
  mutate(
    match_result = factor(
      match_result,
      levels = levels(train_data$match_result)
    ),
    home_confederation = factor(
      home_confederation,
      levels = levels(train_data$home_confederation)
    ),
    away_confederation = factor(
      away_confederation,
      levels = levels(train_data$away_confederation)
    )
  )
```

# Predictor Setup

```{r}
all_predictors <- c(
  "wc_win_rate_diff",
  "wc_gf_pg_diff",
  "wc_ga_pg_diff",
  "wc_appearances_diff",
  "previous_finish_diff",
  "h2h_win_pct",
  "h2h_matches",
  "host_advantage",
  "home_confederation",
  "away_confederation",
  "defending_champion_diff",
  "avg_squad_age_diff",
  "squad_wc_experience_diff",
  "elo_diff",
  "avg_caps_diff",
  "qualification_perf_diff",
  "qualification_gd_diff"
)

numeric_predictors <- setdiff(
  all_predictors,
  c("home_confederation", "away_confederation")
)

full_formula <- reformulate(
  all_predictors,
  response = "match_result"
)

qda_formula <- reformulate(
  numeric_predictors,
  response = "match_result"
)
```

# Standardize Numeric Predictors

Scaling is based only on the training set.

```{r}
train_means <- train_data |>
  summarise(
    across(
      all_of(numeric_predictors),
      mean
    )
  )

train_sds <- train_data |>
  summarise(
    across(
      all_of(numeric_predictors),
      sd
    )
  )

train_scaled <- train_data
test_scaled <- test_data

for (x in numeric_predictors) {
  train_scaled[[x]] <-
    (train_scaled[[x]] - train_means[[x]]) /
    train_sds[[x]]

  test_scaled[[x]] <-
    (test_scaled[[x]] - train_means[[x]]) /
    train_sds[[x]]
}

stopifnot(
  nrow(train_scaled) == 488,
  nrow(test_scaled) == 168,
  sum(is.na(train_scaled)) == 0,
  sum(is.na(test_scaled)) == 0
)
```

# Evaluation Function

```{r}
evaluate_model <- function(
    model_name,
    actual,
    predicted,
    probabilities = NULL) {

  predicted <- factor(
    predicted,
    levels = levels(actual)
  )

  confusion <- table(
    Actual = actual,
    Predicted = predicted
  )

  accuracy <- sum(diag(confusion)) / sum(confusion)

  recall <- diag(confusion) / rowSums(confusion)
  precision <- diag(confusion) / colSums(confusion)

  precision[is.na(precision)] <- 0
  recall[is.na(recall)] <- 0

  f1 <- ifelse(
    precision + recall == 0,
    0,
    2 * precision * recall / (precision + recall)
  )

  macro_f1 <- mean(f1)
  balanced_accuracy <- mean(recall)

  log_loss <- NA_real_

  if (!is.null(probabilities)) {
    probabilities <- as.matrix(probabilities)

    probabilities <- probabilities[
      ,
      levels(actual),
      drop = FALSE
    ]

    actual_column <- match(
      as.character(actual),
      colnames(probabilities)
    )

    actual_probability <- probabilities[
      cbind(
        seq_along(actual),
        actual_column
      )
    ]

    actual_probability <- pmax(
      actual_probability,
      1e-15
    )

    log_loss <- -mean(log(actual_probability))
  }

  list(
    overall = tibble(
      model = model_name,
      accuracy = accuracy,
      macro_f1 = macro_f1,
      balanced_accuracy = balanced_accuracy,
      log_loss = log_loss
    ),

    class = tibble(
      model = model_name,
      class = names(recall),
      precision = as.numeric(precision),
      recall = as.numeric(recall),
      f1 = as.numeric(f1)
    ),

    confusion = as.data.frame(confusion) |>
      as_tibble() |>
      mutate(model = model_name) |>
      relocate(model)
  )
}

results <- list()
```

# Multinomial Logistic Regression

## Full Model

```{r}
full_model <- multinom(
  full_formula,
  data = train_scaled,
  trace = FALSE
)

full_pred <- predict(
  full_model,
  newdata = test_scaled,
  type = "class"
)

full_prob <- predict(
  full_model,
  newdata = test_scaled,
  type = "probs"
)

results[["Full Multinomial"]] <- evaluate_model(
  "Full Multinomial",
  test_scaled$match_result,
  full_pred,
  full_prob
)
```

## Stepwise Model

```{r}
step_model <- MASS::stepAIC(
  full_model,
  direction = "backward",
  trace = FALSE
)

step_pred <- predict(
  step_model,
  newdata = test_scaled,
  type = "class"
)

step_prob <- predict(
  step_model,
  newdata = test_scaled,
  type = "probs"
)

results[["Stepwise Multinomial"]] <- evaluate_model(
  "Stepwise Multinomial",
  test_scaled$match_result,
  step_pred,
  step_prob
)
```

# Discriminant Analysis

## LDA

```{r}
lda_model <- MASS::lda(
  full_formula,
  data = train_scaled
)

lda_output <- predict(
  lda_model,
  newdata = test_scaled
)

lda_pred <- lda_output$class
lda_prob <- lda_output$posterior

results[["LDA"]] <- evaluate_model(
  "LDA",
  test_scaled$match_result,
  lda_pred,
  lda_prob
)
```

## QDA

QDA uses only numeric predictors because sparse factor levels can cause singular class covariance matrices.

```{r}
qda_model <- MASS::qda(
  qda_formula,
  data = train_scaled
)

qda_output <- predict(
  qda_model,
  newdata = test_scaled
)

qda_pred <- qda_output$class
qda_prob <- qda_output$posterior

results[["QDA"]] <- evaluate_model(
  "QDA",
  test_scaled$match_result,
  qda_pred,
  qda_prob
)
```

# Model Matrices and Tournament Folds

```{r}
train_x <- model.matrix(
  ~ . - 1,
  data = train_scaled |>
    select(all_of(all_predictors))
)

test_x <- model.matrix(
  ~ . - 1,
  data = test_scaled |>
    select(all_of(all_predictors))
)

train_y <- train_scaled$match_result
test_y <- test_scaled$match_result

training_years <- sort(
  unique(train_scaled$year)
)

year_fold <- tibble(
  year = training_years,
  fold = rep(
    1:5,
    length.out = length(training_years)
  )
)

fold_id <- year_fold$fold[
  match(
    train_scaled$year,
    year_fold$year
  )
]
```

# K-Nearest Neighbors

```{r}
k_grid <- seq(
  3,
  31,
  by = 2
)

knn_cv_accuracy <- numeric(
  length(k_grid)
)

for (j in seq_along(k_grid)) {
  k <- k_grid[j]
  fold_accuracy <- numeric(5)

  for (fold in 1:5) {
    validation_rows <- which(
      fold_id == fold
    )

    cv_pred <- class::knn(
      train = train_x[-validation_rows, ],
      test = train_x[validation_rows, ],
      cl = train_y[-validation_rows],
      k = k
    )

    fold_accuracy[fold] <- mean(
      cv_pred == train_y[validation_rows]
    )
  }

  knn_cv_accuracy[j] <- mean(
    fold_accuracy
  )
}

knn_cv_results <- tibble(
  k = k_grid,
  cv_accuracy = knn_cv_accuracy
)

best_k <- knn_cv_results |>
  slice_max(
    cv_accuracy,
    n = 1,
    with_ties = FALSE
  ) |>
  pull(k)

knn_prob <- matrix(
  0,
  nrow = nrow(test_x),
  ncol = length(levels(train_y))
)

colnames(knn_prob) <- levels(train_y)

for (i in seq_len(nrow(test_x))) {
  distances <- rowSums(
    sweep(
      train_x,
      2,
      test_x[i, ],
      "-"
    )^2
  )

  neighbors <- order(
    distances
  )[1:best_k]

  neighbor_classes <- factor(
    train_y[neighbors],
    levels = levels(train_y)
  )

  knn_prob[i, ] <-
    as.numeric(
      table(neighbor_classes)
    ) / best_k
}

knn_pred <- factor(
  colnames(knn_prob)[
    max.col(
      knn_prob,
      ties.method = "first"
    )
  ],
  levels = levels(train_y)
)

results[["KNN"]] <- evaluate_model(
  paste0("KNN (k=", best_k, ")"),
  test_y,
  knn_pred,
  knn_prob
)
```

# Decision Tree

```{r}
tree_model <- rpart::rpart(
  full_formula,
  data = train_scaled,
  method = "class",
  control = rpart::rpart.control(
    cp = 0.001,
    xval = 10
  )
)

best_cp <- tree_model$cptable[
  which.min(
    tree_model$cptable[, "xerror"]
  ),
  "CP"
]

tree_model <- rpart::prune(
  tree_model,
  cp = best_cp
)

tree_pred <- predict(
  tree_model,
  newdata = test_scaled,
  type = "class"
)

tree_prob <- predict(
  tree_model,
  newdata = test_scaled,
  type = "prob"
)

results[["Decision Tree"]] <- evaluate_model(
  "Decision Tree",
  test_y,
  tree_pred,
  tree_prob
)
```

# Random Forest

```{r}
set.seed(131)

rf_model <- randomForest::randomForest(
  full_formula,
  data = train_scaled,
  ntree = 500,
  importance = TRUE
)

rf_pred <- predict(
  rf_model,
  newdata = test_scaled,
  type = "response"
)

rf_prob <- predict(
  rf_model,
  newdata = test_scaled,
  type = "prob"
)

results[["Random Forest"]] <- evaluate_model(
  "Random Forest",
  test_y,
  rf_pred,
  rf_prob
)
```

# XGBoost

```{r}
set.seed(131)

xgb_y <- as.integer(train_y) - 1

dtrain <- xgboost::xgb.DMatrix(
  data = train_x,
  label = xgb_y
)

xgb_params <- list(
  objective = "multi:softprob",
  num_class = 3,
  eval_metric = "mlogloss",
  max_depth = 4,
  eta = 0.05,
  subsample = 0.8,
  colsample_bytree = 0.8
)

xgb_folds <- lapply(
  sort(unique(fold_id)),
  function(f) {
    which(fold_id == f)
  }
)

xgb_cv <- xgboost::xgb.cv(
  params = xgb_params,
  data = dtrain,
  nrounds = 500,
  folds = xgb_folds,
  early_stopping_rounds = 25,
  verbose = 0
)

best_rounds <- xgb_cv$early_stop$best_iteration

if (
  is.null(best_rounds) ||
  length(best_rounds) == 0
) {
  best_rounds <- which.min(
    xgb_cv$evaluation_log$test_mlogloss_mean
  )
}

xgb_model <- xgboost::xgb.train(
  params = xgb_params,
  data = dtrain,
  nrounds = best_rounds,
  verbose = 0
)

raw_prob <- predict(
  xgb_model,
  newdata = test_x
)

if (is.matrix(raw_prob)) {
  xgb_prob <- raw_prob
} else {
  xgb_prob <- matrix(
    raw_prob,
    ncol = 3,
    byrow = TRUE
  )
}

colnames(xgb_prob) <- levels(test_y)

xgb_pred <- factor(
  colnames(xgb_prob)[
    max.col(
      xgb_prob,
      ties.method = "first"
    )
  ],
  levels = levels(test_y)
)

results[["XGBoost"]] <- evaluate_model(
  "XGBoost",
  test_y,
  xgb_pred,
  xgb_prob
)
```

# Regularized Multinomial Regression

## Ridge

```{r}
set.seed(131)

ridge_cv <- glmnet::cv.glmnet(
  x = train_x,
  y = train_y,
  family = "multinomial",
  alpha = 0,
  foldid = fold_id,
  type.measure = "deviance"
)

ridge_lambda <- ridge_cv$lambda.min

ridge_model <- glmnet::glmnet(
  x = train_x,
  y = train_y,
  family = "multinomial",
  alpha = 0,
  lambda = ridge_lambda
)

ridge_prob <- predict(
  ridge_model,
  newx = test_x,
  s = ridge_lambda,
  type = "response"
)[, , 1]

colnames(ridge_prob) <- levels(train_y)

ridge_pred <- factor(
  colnames(ridge_prob)[
    max.col(
      ridge_prob,
      ties.method = "first"
    )
  ],
  levels = levels(test_y)
)

results[["Ridge Multinomial"]] <- evaluate_model(
  "Ridge Multinomial",
  test_y,
  ridge_pred,
  ridge_prob
)
```

## Elastic Net

```{r}
set.seed(131)

elastic_alpha_grid <- c(
  0.25,
  0.50,
  0.75
)

elastic_cv_results <- map_dfr(
  elastic_alpha_grid,
  function(a) {
    fit <- glmnet::cv.glmnet(
      x = train_x,
      y = train_y,
      family = "multinomial",
      alpha = a,
      foldid = fold_id,
      type.measure = "deviance"
    )

    tibble(
      alpha = a,
      lambda = fit$lambda.min,
      cv_deviance = min(fit$cvm)
    )
  }
)

best_elastic <- elastic_cv_results |>
  slice_min(
    cv_deviance,
    n = 1,
    with_ties = FALSE
  )

best_elastic_alpha <- best_elastic$alpha
best_elastic_lambda <- best_elastic$lambda

elastic_model <- glmnet::glmnet(
  x = train_x,
  y = train_y,
  family = "multinomial",
  alpha = best_elastic_alpha,
  lambda = best_elastic_lambda
)

elastic_prob <- predict(
  elastic_model,
  newx = test_x,
  s = best_elastic_lambda,
  type = "response"
)[, , 1]

colnames(elastic_prob) <- levels(train_y)

elastic_pred <- factor(
  colnames(elastic_prob)[
    max.col(
      elastic_prob,
      ties.method = "first"
    )
  ],
  levels = levels(test_y)
)

results[["Elastic Net Multinomial"]] <- evaluate_model(
  "Elastic Net Multinomial",
  test_y,
  elastic_pred,
  elastic_prob
)

elastic_coef <- coef(
  elastic_model,
  s = best_elastic_lambda
)

elastic_coef_table <- bind_rows(
  lapply(
    names(elastic_coef),
    function(class_name) {
      mat <- as.matrix(
        elastic_coef[[class_name]]
      )

      tibble(
        class = class_name,
        predictor = rownames(mat),
        coefficient = as.numeric(mat[, 1])
      )
    }
  )
) |>
  arrange(
    class,
    desc(abs(coefficient))
  )
```

# Majority-Class Baseline

```{r}
class_proportions <- prop.table(
  table(train_y)
)

baseline_class <- names(
  which.max(class_proportions)
)

baseline_pred <- factor(
  rep(
    baseline_class,
    length(test_y)
  ),
  levels = levels(test_y)
)

baseline_prob <- matrix(
  rep(
    as.numeric(class_proportions),
    each = length(test_y)
  ),
  nrow = length(test_y)
)

colnames(baseline_prob) <- names(
  class_proportions
)

results[["Baseline"]] <- evaluate_model(
  "Majority Baseline",
  test_y,
  baseline_pred,
  baseline_prob
)
```

# Combine Results

```{r}
overall_results <- bind_rows(
  lapply(
    results,
    function(x) x$overall
  )
) |>
  arrange(
    desc(accuracy)
  )

class_results <- bind_rows(
  lapply(
    results,
    function(x) x$class
  )
)

confusion_results <- bind_rows(
  lapply(
    results,
    function(x) x$confusion
  )
)

multinomial_aic <- tibble(
  model = c(
    "Full Multinomial",
    "Stepwise Multinomial"
  ),
  AIC = c(
    AIC(full_model),
    AIC(step_model)
  )
)

test_predictions <- test_scaled |>
  transmute(
    year,
    match_id,
    home_team_name,
    away_team_name,
    actual = match_result,
    full_multinomial = full_pred,
    stepwise_multinomial = step_pred,
    lda = lda_pred,
    qda = qda_pred,
    knn = knn_pred,
    decision_tree = tree_pred,
    random_forest = rf_pred,
    xgboost = xgb_pred,
    ridge_multinomial = ridge_pred,
    elastic_net_multinomial = elastic_pred
  )
```

# Variable Importance

```{r}
rf_importance <- randomForest::importance(
  rf_model
) |>
  as.data.frame() |>
  rownames_to_column("predictor")

xgb_importance <- xgboost::xgb.importance(
  model = xgb_model
)
```

# Save Results

```{r}
dir.create(
  "data",
  showWarnings = FALSE
)

dir.create(
  "models",
  showWarnings = FALSE
)

write_csv(
  overall_results,
  "data/model_comparison_results.csv"
)

write_csv(
  class_results,
  "data/model_class_metrics.csv"
)

write_csv(
  confusion_results,
  "data/model_confusion_matrices.csv"
)

write_csv(
  multinomial_aic,
  "data/multinomial_aic_results.csv"
)

write_csv(
  knn_cv_results,
  "data/knn_cv_results.csv"
)

write_csv(
  elastic_cv_results,
  "data/elastic_net_cv_results.csv"
)

write_csv(
  elastic_coef_table,
  "data/elastic_net_coefficients.csv"
)

write_csv(
  test_predictions,
  "data/model_test_predictions.csv"
)

write_csv(
  rf_importance,
  "data/random_forest_importance.csv"
)

write_csv(
  xgb_importance,
  "data/xgboost_importance.csv"
)

saveRDS(
  full_model,
  "models/full_multinomial.rds"
)

saveRDS(
  step_model,
  "models/stepwise_multinomial.rds"
)

saveRDS(
  lda_model,
  "models/lda_model.rds"
)

saveRDS(
  qda_model,
  "models/qda_model.rds"
)

saveRDS(
  tree_model,
  "models/decision_tree.rds"
)

saveRDS(
  rf_model,
  "models/random_forest.rds"
)

saveRDS(
  xgb_model,
  "models/xgboost.rds"
)

saveRDS(
  ridge_model,
  "models/ridge_multinomial.rds"
)

saveRDS(
  elastic_model,
  "models/elastic_net_multinomial.rds"
)

saveRDS(
  list(best_k = best_k),
  "models/knn_settings.rds"
)
```

# Final Results

```{r}
overall_results

multinomial_aic

formula(step_model)

knn_cv_results

elastic_cv_results

elastic_coef_table

cat(
  "\nBest KNN k:",
  best_k,
  "\n"
)

cat(
  "Best XGBoost rounds:",
  best_rounds,
  "\n"
)

cat(
  "Ridge lambda:",
  ridge_lambda,
  "\n"
)

cat(
  "Best Elastic Net alpha:",
  best_elastic_alpha,
  "\n"
)

cat(
  "Best Elastic Net lambda:",
  best_elastic_lambda,
  "\n"
)
```
